\textsc{TheHaloMod}: An online calculator for the halo model

### 2020-09-28 paper.tex
Almost ready for submission

\listfiles
\documentclass[5p]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{xcolor}
\usepackage{longtable,booktabs,tabu}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage{pbox}
\usepackage{amssymb}

@@

\newcommand{\Nc}{\langle N_c \rangle}
\newcommand{\Ns}{\langle N_s \rangle}
\newcommand{\Nt}{\langle N_t \rangle}
\newcommand{\halomod}{\textsc{halomod}}
\newcommand{\thm}{\textsc{TheHaloMod}}
\newcommand{\hmf}{\textsc{hmf}}

\newcommand{\bd}[1]{\textcolor{purple}{\textbf{[BD: #1]}}}
\newcommand{\bde}[1]{\textcolor{blue}{#1}}
\newcommand{\sgm}[1]{\textcolor{green}{\textbf{[SM: #1]}}}

@@

%% Group authors per affiliation:
\author{Steven~G.~Murray}
\address{Arizona State University, Tempe, AZ, USA}

@@

We use this framework to motivate the presentation of a new Python tool for simple and efficient calculation of halo model quantities, called \halomod. This tool is efficient, simple to use, comprehensive and importantly provides a great deal of flexibility in terms of custom extensions. These characteristics will accommodate novel developments for years to come.
This Python tool is complemented by a new web-application that supports generation of many halo model quantities directly from the browser -- useful for educators, students, theorists and observers.

@@

\begin{keyword}

\end{keyword}

@@

@@

In practice, the HOD formalism has been widely used in the interpretation of galaxy populations in the past decade. Most of these studies have focused on determining the parameters of the HOD (i.e. the galaxy-halo connection) from the 2PCF of galaxies \citep[eg.][]{Moustakas2002,Bullock2002,Zheng2004,Zehavi2005,Blake2008,Zehavi2011,Beutler2013,Skibba2015} using the analytical framework that we present in this paper. However, other observables such as galaxy-galaxy lensing \citep{Mandelbaum2006}, the 2PCF of radio galaxies \citep{Wake2008a,Kim2011}, galaxy-quasar cross-correlations \citep{Shen2013},near-UV cross-correlations \citep{Krause2012} and \textsc{Hi} intensity map cross-correlations \citep{Padmanabhan2016b,Wolz2019} have also received application through the same framework. Furthermore, several studies have successfully combined observables to break degeneracies \citep{Leauthaud2011,Leauthaud2012,More2013}, and the combination of the galaxy 2PCF with weak-lensing \citep{More2015}, stellar mass functions \citep{Coupon2015} and group mass-to-number ratios \citep{Reddick2014} has resulted in the ability to simultaneously constrain cosmological parameters along with those of the HOD.
%Several related approaches that employ different means for connecting galaxies to halos have been developed, including conditional luminosity functions \citep[CLF;][]{Yang2003,vandenBosch2003,Yang2005,Cooray2005,
%Cooray2005a,Cooray2006,vandenBosch2013,Cacciato2013} and subhalo abundance matching (SHAM; eg. \citet{Conroy2006}). \bd{This paragraph sort of comes out of nowhere, and feels a bit disconnected. Also, given the long refernece lists for HOD models, just citing one paper for SHAM might piss some people off... I'd suggest we take it out altogether. Then, the next paragraph properly connects to the previous one as well. Moreover, I don't think SHAM is ever applied to halo-model ``data''? I've only ever seen it applied to simulated catalogs.}

Clearly the halo model (hereafter HM) framework, complemented by an HOD or comparable mechanism, can be of wide utility in the interpretation of large surveys.
Increasingly, observational studies are employing a non-analytic incarnation of the HM, in which simulated halos are ``painted'' with a particular galaxy sample, given detailed semi-analytic models of the galaxy-halo connection \citep{XXX}.
In particular, the \textsc{halotools}\footnote{\url{https://halotools.readthedocs.io}} library has become a popular implementation of this approach.
Nevertheless, a purely analytic construction of the HM is still of great importance and utility: it is our best fundamental model of the non-linear scales of the matter distribution.

The HM is a complex framework: it synthesises many related sub-components (eg. halo profiles, mass functions, bias models, spatial filters, halo exclusion models, concentration-mass relations...) to produce spatial statistics. 
These sub-components can often be modelled independently via simulations, and new more accurate models are being produced regularly by the community. 
This highlights the need for an implementation of the HM framework that has the flexibility and modularity to enable easy switching between models for the various sub-components, and rapid development of new models to incorporate into the framework.

@@

%However, while the HM has become an increasingly popular tool, it is still difficult to find a publicly-available, stable and efficient code for its calculation. \bd{I think this will annoy anyone who has written such a code because those are very fundamental benchmarks. Maybe we could say that those codes are not fully flexible?} 
%To be clear, it is not that implementations do not exist 
While there are a few publicly available implementations of the HM (many of which  were used in the testing of our code, though most have discontinued development),
eg. \textsc{chomp}\footnote{\url{https://code.google.com/p/chomp/}  -- discontinued.}, \textsc{HMcode}\footnote{\url{https://github.com/alexander-mead/HMcode/}} and \textsc{AUM}\footnote{\url{https://github.com/surhudm/aum} -- discontinued.},
%, nor that they are poor implementations in general (indeed, all of these were useful in testing our own code). 
it is our experience from that a remarkable number of 
%researchers are using the HM in various ways, but almost ubiquitously they find need to write their own code -- 
practitioners develop their own tools --- whether based heavily on existing (public or private) code or from the ground up.

@@

    \item \textbf{Flexible/Extendible. } The HM is a rapidly evolving framework, with individual components constantly improving, and the framework itself being extended. Building a static implementation is therefore non-conducive  to the development of the field. Components need to be as plug-and-play as possible, with new models easily created and inserted on the fly. Our implementation of 
    such a plug-and-play system is outlined in \S\ref{sec:halomod:overview:flexibility}.

@@

Both philosophically and technically, \halomod\ inherits from the \hmf\ halo mass function package\footnote{https://github.com/steven-murray/hmf} which was first presented in \cite{Murray2013a}.
Many developments have occurred in \hmf\ since its first publication, and the technical framework of \halomod\ presented in this paper is essentially inherited from the updates in \hmf. 
Thus, this paper can also secondarily be considered an update of \hmf.

@@

In addition to \halomod\ and \hmf\ package, we also present a new web-application, \thm\, which is able to generate full halo model quantities (eg. two-point correlation functions and galaxy power spectra) without ever having to install the Python package.
It is a successor to the popular \textsc{HMFcalc} \citep{Murray2013a} web-application, and includes the full range of functionality of \textsc{HMFcalc}.
The presentation of \thm\ completes our vision for \halomod\ with (v) a tool for educators to easily and interactively present cosmologically relevant quantities graphically.

@@

@@

The application of the HM rests in converting the premise of Eq. \ref{eq:hm_sum} into a semi-analytic integration. This requires knowledge of four key components:
\begin{enumerate}
    \item The average radial density profile of halos, $\rho(r|m)$
    \item The abundance of halos of a given mass (termed the halo mass function, or HMF), $n(m)$
    \item The expected overdensity of halos of mass $m$, $\delta_h(\mathbf{x},m)$ in a region with a given overdensity of matter $\delta(\mathbf{x})$, called the \textit{halo bias}.
    \item A model for the spatial distribution of matter from a linea 
\end{enumerate}
In practice, considering two-point clustering also requires a knowledge of the spatial clustering of halos, most commonly specified through the spatial clustering of matter.\bd{I was missing that from the list. Is there any alternative?} Furthermore, detailed calculations require the following extra components:
\begin{enumerate}
    \setcounter{enumi}{3}
    \item Additional parameters that describe the halo profile, most
    commonly parameterized as a concentration-mass relation, $c(m)$, linking the shape of a halo profile to its mass.
    \item A model for ``halo exclusion", which accounts for double-counting intra-halo correlations.
    \item To extend the calculations to galaxies, a distribution function for the occupation of halos by galaxies, as a function of halo mass, $N(m)$ (the HOD; cf. \S\ref{sec:TracerHaloModel}). 
\end{enumerate}

@@

\subsubsection{2-halo term}
\label{sec:theory:2halo}
The two-halo term is most elegantly expressed in Fourier-space, \bd{the notation is different from the 1-halo term, where it's $P(k)$, here it's $P(k,r)$. Why? Might be good to explain why the scale-dependence falls out of the 1-halo temr}
\sgm{So this is interesting. There's an $r$ in the 2-halo term because either the halo exclusion equations include halo radius scales, or the halo bias is ``scale-dependent'' which may be defined in real-space. I deal with this by hankel transforming to $\xi(r)$ and then back to $P(k)$. Is that right?}
\sgm{So just write a sentence explaining this.}
\begin{align}
    P_{\rm DM}^{2h}(k,r) =  \int \int & dm_1 dm_2\  I_{\rm DM}(k,m_1) I_{\rm DM}(k,m_2) \nonumber \\ 
    &\times P_{hh}(k,r|m_1,m_2),
\end{align}
where
\begin{equation}
    I_{\rm DM}(k,m) = n(m) u(k|m) \frac{m}{\bar{\rho}}.
\end{equation}
Here, $P_{hh}(k,r|m_1,m_2)$ is the power spectrum of halo centres for halos of mass $m_1$ and $m_2$. This is in general a complicated function of mass and scale, but the most common approach is to approximate it by a first-order linear bias of the matter power,
\begin{equation}
    \label{eq:halo_centre_power}
    P_{hh}(k,r|m_1,m_2) \approx b(m_1,r)b(m_2,r)P_m(k),
\end{equation}
where $b(m,r)$ is the first-order bias at $m$ (cf. \S\ref{sec:biastheory}), possibly with a scale-dependent correction, and $P_m(k)$ is the matter power spectrum (cf. \S\ref{sec:powerspec})\footnote{Note that thoughout, we use a subscript $m$ to denote either the linear matter power or the nonlinear matter power derived using analytic corrections, as opposed to the full halo model.}. Typically, empirical nonlinear corrections on the linear matter power spectrum from \textsc{halofit} \citep{Smith2003} are used for this quantity. We leave its definition free in our implementation. \bd{this seems circular: doesn't Smith+03 use the halo model to derive their corrections?} \sgm{Yup. It confuses me too, but it's what has been done in the literature, eg. Smith and Markovic 2011.}

The general form for the 2-halo term then becomes
\begin{align}
\label{eq:dmpower2a}
    P_{\rm DM}^{2h}(k,r) =  \int\int & I_{\rm DM}(k,m_1) I_{\rm DM}(k,m_2) \nonumber \\
    &\times b(m_1,r)b(m_2,r)P_{m}(k) dm_2 dm_1
\end{align}
This is separable, so long as the integration limits are not inter-dependent (this is not the case in general, cf. \S\ref{sec:theory:exclusion}), and in this simplest case \bd{what case is that? no correlations? In what case can we not combine the $m_1$ and $m_2$ integrals?} \sgm{Any case that doesn't try to exclude halos.} we have
\begin{equation}
    \label{eq:dmpower2}
    P_{DM}^{2h}(k,r) = P_m(k) \left[ \int  I_\text{DM}(k,m) b(m,r) dm \right]^2.
\end{equation}

@@

The CDM transfer function has been modelled extensively in the literature. 
%At least three 
\bde{A number of } functional forms have been proposed \citep[][hereafter EH]{Bond1984,Bardeen1986,Eisenstein1999}, along with a series of \textit{Boltzmann} codes, which calculate the transfer function from first principles with perturbation theory techniques \citep{Zaldarriaga2000,Lewis2000,Blas2011}.

Figure \ref{fig:power} shows the typical shape of the CDM linear power spectrum, while highlighting differences between forms for $T(k)$ found in the literature. Both small- and large-scales have small fluctuations, with a characteristic bend at scales of $\sim 0.02h/$Mpc. Only the Boltzmann code is able to fully capture the wiggles introduced by the baryon acoustic oscillations (BAOs), though the EH fit closely follows in all other respects. In fact, all curves (except that of \citet{Bond1984}) are within $\sim 10\%$ over the plotted range.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/power_fits.pdf}
  \caption[Comparison of transfer function models]{A selection of transfer function models, including analytic approximations \citep{Bond1984,Bardeen1986,Eisenstein1998} and Boltzmann solutions. The general shape is reproduced in each case, with details of baryon acoustic oscillations differing. In the bottom panel, we show the relative fraction compared to the \citet{Eisenstein1999} transfer functions without baryons. Also plotted are two versions of the empirical nonlinear corrections which lead to enhanced power on small scales. \bd{Figure comments: frame around legend cuts into lines, could be removed; why are non-linear versions not compared in bottom panel?; some labels could be made prettier, e.g. BondEfs to Bond\& Efstathiou; tick labels are very small, bordering on illegible; I think it would make more sense to compare the fits to CAMB, i.e., the closest to the true TF?}}
  \label{fig:power}
\end{figure}

The calculation of the halo-centre power spectrum \bd{not sure what that term means? The halo-halo PS?} \sgm{yeah, halo-halo. Some papers call it the halo-centre power} is commonly approximated by Eq. \ref{eq:halo_centre_power}, i.e.
\begin{equation*}
    P_{hh}(k,r|m_1,m_2) \approx b(m_1,r)b(m_2,r)P_m(k),
\end{equation*}
 in which an estimate of the matter power spectrum is linearly biased. Typically, a \textit{nonlinear} estimate of the matter power is used, for which the corrections of \citet[][hereafter \textsc{halofit}]{Smith2003} are employed. We do not present the details here, except to mention that the modifications are applied as ratios of $P_\text{m,lin}(k)$, which are obtained via arguments motivated by the halo model. These modifications are expressed as a series of equations with coefficients fitted to numerical simulations. Updated coefficients were introduced in \citet{Takahashi2012}. Note also that the normalisation of $P_\text{m,nl}(k)$ is identical to $P_\text{m,lin}(k)$, and is calculated using the linear form, since $\sigma_8$ refers to the linear power spectrum. The effect of introducing non-linearities is to increase small-scale power (cf. figure \ref{fig:power}), with the recent coefficient updates slightly enhancing this effect.

@@

The SO algorithm finds peaks in the density field and grows spheres around them until the contained density drops below a threshold density.
All particles within the final sphere are considered part of the halo. 
This definition is more useful in the context of observations (of clusters, for example) as it can be applied to the measured (projected) density fields. \bd{Sort of disagree... SO masses are still never measured. It's true that they can be projected in theory, but you still need to forward-model observations and use scaling relations} \sgm{True, but in comparison to FoF, they are still more useful observationally, right?}

@@

Having defined a spatial filter on the primordial density field, one can proceed to define the Lagrangian radius, $R(m)$, which is the radius of a spherical region of homogeneous space containing the mass $m$ (or more generally, the scale of a particular filter function required to contain the mass, cf. \S\ref{sec:halomod:components:window}). \bd{Physically, what this means is that we consider the variance on a scale from which the matter in halos is thought to have originated.} For the standard top-hat, this is
\begin{equation}
    R_L(m) = \left(\frac{3m}{4\pi\bar{\rho}_0}\right)^{1 / 3},
\end{equation}
with $\bar{\rho}_0$ the mean density of the Universe at the present epoch.

The ``mass variance'', $\sigma^2(R_L)$ is thus the variance on a scale from which the matter in halos is though to have originated. Notably, the mass variance in spheres of Lagrangian radius 8 Mpc/$h$ yields the power spectrum normalisation, $\sigma_8$.

\subsection{The Mass Function}
\label{sec:massfunction}
The mass function quantifies the number of dark matter halos per unit
mass per unit comoving volume of the Universe. 
%We do not include a detailed summary of the theory of the mass function here, and direct the reader to \citet{Murray2013a} for a review and table of fitting functions. 
The most commonly adopted theoretical formulation of the mass function is via the extended Press-Schechter (EPS) formalism \citep{Press1974,Bond1991}, in which density peaks form through gravitational instability and evolve into halos when they cross a density threshold, $\delta_c$. 
In this formalism, the mass function is a universal function of the mass variance, $\sigma$, which captures the variation from cosmological parameters and the redshift \citep[but note that several recent studies find significant non-universality with redshift, and have provided fits that explicitly account for this departure]{Jenkins2001}. \bd{NOT TRUE! I thought that too until recently... Andrey corrected me. The PS mass function is universal only if the collapse threshold is universal, i.e. $\delta_c(\nu)$; the excursion set models are generally universal only for a sharp-k filter I believe. So universality is an ASSUMPTION, not a result of these models. Again, discussed in detail in my MF paper.} 
\sgm{Come back to this.}

>>>>

\listfiles
\documentclass[5p,aas_macros]{elsarticle}

\usepackage{xcolor}
\usepackage{lineno}
\usepackage[colorlinks=true,urlcolor=black,linkcolor=blue,citecolor=blue]{hyperref}
\usepackage{minted}
\usepackage{longtable,booktabs,tabu}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage{pbox}
\usepackage{amssymb}
\usepackage{layouts}
\usepackage{subfig}
\usepackage{microtype}
%\usepackage{subcaption}
%\usepackage{graphicx}

\input{citation_fix}

@@

\newcommand{\Nc}{\langle N_c \rangle}
\newcommand{\Ns}{\langle N_s \rangle}
\newcommand{\Nt}{\langle N_t \rangle}
\newcommand{\halomod}{\textsc{halomod}}
\newcommand{\thm}{\textsc{TheHaloMod}}
\newcommand{\hmf}{\textsc{hmf}}
\newcommand{\framework}{\texttt{Framework}}
\newcommand{\component}{\texttt{Component}}
\newcommand{\parameter}{\texttt{parameter}}
\newcommand{\cached}{\texttt{cached\_quantity}}

\newcommand{\bd}[1]{\textcolor{purple}{\textbf{[BD: #1]}}}
\newcommand{\bde}[1]{\textcolor{blue}{#1}}
\newcommand{\sgm}[1]{\textcolor{green}{\textbf{[SM: #1]}}}
\newcommand{\ztc}[1]{\textcolor{cyan}{\textbf{[ZTC: #1]}}}

\def\apj{ApJ}
\def\mnras{MNRAS}

\usemintedstyle{friendly}

\setminted[python]{
    frame=leftline,
    framesep=2mm,
    baselinestretch=1.2,
    %bgcolor=lightgray,
    fontsize=\small,
}

@@

%% Group authors per affiliation:
\author{Steven~G.~Murray$^a$, Benedikt Diemer$^b$, Zhaoting Chen$^c$}
\address{$^a$ Arizona State University, Tempe, AZ, USA \\
$^b$ Department of Astronomy, University of Maryland, College Park, MD 20742, USA\\
$^c$ Jodrell Bank Centre for Astrophysics, School of Physics and Astronomy, The University of Manchester, Manchester M13 9PL, UK}

@@

We use this framework to motivate the presentation of a new Python tool for simple and efficient calculation of halo model quantities, and their extension to galaxy statistics via a \textit{halo occupation distribution}, called \halomod. This tool is efficient, simple to use, comprehensive and importantly provides a great deal of flexibility in terms of custom extensions. 
This Python tool is complemented by a new web-application at \url{https://thehalomod.app} that supports the generation of many halo model quantities directly from the browser -- useful for educators, students, theorists and observers.

@@

\begin{keyword}
large-scale structure of universe --
dark matter --
galaxies: halos --
methods: analytical --
methods: numerical
\end{keyword}

@@

%\printinunitsof{in}\prntlen{\textwidth}
%\printinunitsof{in}\prntlen{\linewidth}

@@

In practice, the HOD formalism has been widely used in the interpretation of galaxy populations in the past decade. Most of these studies have focused on determining the parameters of the HOD (i.e. the galaxy-halo connection) from the 2PCF of galaxies \citep[eg.][]{Moustakas2002,Bullock2002,Zheng2004,Zehavi2005,Blake2008,Zehavi2011,Beutler2013,Skibba2015} using the analytical framework that we present in this paper. However, other observables such as galaxy-galaxy lensing \citep{Mandelbaum2006}, the 2PCF of radio galaxies \citep{Wake2008a,Kim2011}, galaxy-quasar cross-correlations \citep{Shen2013}, near-UV cross-correlations \citep{Krause2012} and \textsc{Hi} intensity map cross-correlations \citep{Padmanabhan2016b,Wolz2019} have also received application through the same framework. Furthermore, several studies have successfully combined observables to break degeneracies \citep{Leauthaud2011,Leauthaud2012,More2013}, and the combination of the galaxy 2PCF with weak-lensing \citep{More2015}, stellar mass functions \citep{Coupon2015} and group mass-to-number ratios \citep{Reddick2014} has resulted in the ability to simultaneously constrain cosmological parameters along with those of the HOD.
%Several related approaches that employ different means for connecting galaxies to halos have been developed, including conditional luminosity functions \citep[CLF;][]{Yang2003,vandenBosch2003,Yang2005,Cooray2005,
%Cooray2005a,Cooray2006,vandenBosch2013,Cacciato2013} and subhalo abundance matching (SHAM; eg. \citet{Conroy2006}). \bd{This paragraph sort of comes out of nowhere, and feels a bit disconnected. Also, given the long refernece lists for HOD models, just citing one paper for SHAM might piss some people off... I'd suggest we take it out altogether. Then, the next paragraph properly connects to the previous one as well. Moreover, I don't think SHAM is ever applied to halo-model ``data''? I've only ever seen it applied to simulated catalogs.}

Clearly, the halo model (hereafter HM) framework, complemented by an HOD or comparable mechanism, can be of wide utility in the interpretation of large surveys.
Increasingly, observational studies are employing a non-analytic incarnation of the HM, in which simulated halos are ``painted'' with a particular galaxy sample, given detailed semi-analytic models of the galaxy-halo connection \citep[eg.][]{Carretero2015}.
In particular, the \textsc{halotools}\footnote{\url{https://halotools.readthedocs.io}} library has become a popular implementation of this approach.
Nevertheless, a purely analytic construction of the HM is still of great importance and utility: it is our best fundamental model of the non-linear scales of the matter distribution.

The HM is a complex framework as it synthesises many related sub-components (eg. halo profiles, mass functions, bias models, spatial filters, halo exclusion models, concen\-tration--mass relations) to produce spatial statistics. 
These sub-components can often be modelled independently via simulations, and new more accurate models are being produced regularly by the community. 
This highlights the need for an implementation of the HM framework that has the flexibility and modularity to enable easy switching between models for the various sub-components, and rapid development of new models to incorporate into the framework.

@@

%However, while the HM has become an increasingly popular tool, it is still difficult to find a publicly-available, stable and efficient code for its calculation. \bd{I think this will annoy anyone who has written such a code because those are very fundamental benchmarks. Maybe we could say that those codes are not fully flexible?} 
%To be clear, it is not that implementations do not exist 
There are a few publicly available implementations of the HM, many of which were used in the testing of our code although the development of most of them has been discontinued (e.g., \textsc{chomp}\footnote{\url{https://code.google.com/p/chomp/}  -- discontinued.}, \textsc{HMcode}\footnote{\url{https://github.com/alexander-mead/HMcode/}} and \textsc{AUM}\footnote{\url{https://github.com/surhudm/aum} -- discontinued.}).
%, nor that they are poor implementations in general (indeed, all of these were useful in testing our own code). 
However, it is our experience from that a remarkable number of 
%researchers are using the HM in various ways, but almost ubiquitously they find need to write their own code -- 
practitioners develop their own tools --- whether based heavily on existing (public or private) code or from the ground up.

@@

    \item \textbf{Flexible/Extendible. } The HM is a rapidly evolving framework, with individual components const\-antly improving, and the framework itself being extended. Building a static implementation is therefore non-conducive  to the development of the field. Components need to be as plug-and-play as possible, with new models easily created and inserted on the fly. Our implementation of 
    such a plug-and-play system is outlined in \S\ref{sec:halomod:overview:flexibility}.

@@

Both philosophically and technically, \halomod\ inherits from the \hmf\ halo mass function package\footnote{\url{https://github.com/steven-murray/hmf}} which was first presented in \cite{Murray2013a}.
Many developments have occurred in \hmf\ since its first publication, and the technical framework of \halomod\ presented in this paper is essentially inherited from the updates in \hmf. 
Thus, this paper can also secondarily be considered an update of \hmf.

@@

In addition to \halomod\ and \hmf\ package, we also present a new web-application, \thm\footnote{\url{ https://thehalomod.app}}, which is able to generate full halo model quantities (eg. two-point correlation functions and galaxy power spectra) without ever having to install the Python package.
It is a successor to the popular \textsc{HMFcalc} \citep{Murray2013a} web-application, and includes the full range of functionality of \textsc{HMFcalc}.
The presentation of \thm\ completes our vision for \halomod\ with (v) a tool for educators to easily and interactively present cosmologically relevant quantities graphically.

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{figures/thehalomod_app_schematic.pdf} 
    \caption{\textsc{TheHaloMod} in action. The main page (right panel) shows the currently computed models, with a selection of quantities to plot, and a number of actions available for each model (edit/clone/report bug/delete). The input form (two insets on the left) provides a wide range of options, including all of the components presented throughout this paper. Each component is defined in its own tab. Those shown here are cosmology and the HOD model.} \label{fig:webpage-image}
\end{figure*}

@@

Note that the code to produce all figures of different component models in this paper, as well as the example application, are available publicly as examples in \halomod's documentation\footnote{At \url{https://halomod.readthedocs.io/en/latest/examples/component-showcase.html} and \url{https://halomod.readthedocs.io/en/latest/examples/fitting.html} respectively.}. This paper refers to \textsc{hmf} version 3.2.1 and \textsc{halomod} version 2.0.0.

@@

The application of the HM rests in converting the prem\-ise of Eq. \ref{eq:hm_sum} into a semi-analytic integration. This requires knowledge of four key components:
\begin{enumerate}
    \item The average radial density profile of halos, $\rho(r|m)$
    \item The abundance of halos of a given mass (termed the halo mass function, or HMF), $n(m)$
    \item The expected overdensity of halos of mass $m$, $\delta_h(\mathbf{x},m)$ in a region with a given overdensity of matter $\delta(\mathbf{x})$, called the \textit{halo bias}.
    \item A model for the linear spatial distribution of matter, $\delta(\mathbf{x})$.
\end{enumerate}
Furthermore, detailed calculations require the following extra components:
\begin{enumerate}
    \setcounter{enumi}{3}
    \item Additional parameters that describe the halo profile, most
    commonly parameterized as a concentration-mass relation, $c(m)$, linking the shape of a halo profile to its mass.
    \item A model for ``halo exclusion", which accounts for double-counting intra-halo correlations.
    \item To extend the calculations to galaxies, a distribution function for the occupation of halos by galaxies, as a function of halo mass, $N(m)$ (the HOD; cf. \S\ref{sec:TracerHaloModel}). 
\end{enumerate}

@@

\subsubsection{2-halo term}
\label{sec:theory:2halo}
The two-halo term is most elegantly expressed in Fourier-space, 
\begin{align}
    P_{\rm DM}^{2h}(k,r) =  \int \int & dm_1 dm_2\  I_{\rm DM}(k,m_1) I_{\rm DM}(k,m_2) \nonumber \\ 
    &\times P_{hh}(k,r|m_1,m_2),
\end{align}
where
\begin{equation}
    I_{\rm DM}(k,m) = n(m) u(k|m) \frac{m}{\bar{\rho}}.
\end{equation}
Here, $P_{hh}(k,r|m_1,m_2)$ is the power spectrum of halo centres for halos of mass $m_1$ and $m_2$. This is in general a complicated function of mass and scale, but the most common approach is to approximate it by a first-order linear bias of the matter power,
\begin{equation}
    \label{eq:halo_centre_power}
    P_{hh}(k,r|m_1,m_2) \approx b(m_1,r)b(m_2,r)P_m(k),
\end{equation}
where $b(m,r)$ is the first-order bias at $m$ (cf. \S\ref{sec:biastheory}), possibly with a scale-dependent correction, and $P_m(k)$ is the matter power spectrum (cf. \S\ref{sec:powerspec})\footnote{Note that thoughout, we use a subscript $m$ to denote either the linear matter power or the nonlinear matter power derived using analytic corrections, as opposed to the full halo model.}. Typically, empirical nonlinear corrections on the linear matter power spectrum from \textsc{halofit} \citep{Smith2003} are used for this quantity. We leave its definition free in our implementation\footnote{This is slightly circular, as the \textsc{halofit} model itself uses the halo model. However, this is a common practice in halo model calculations, eg. \citep{Cooray2002, Smith2011}}.

The general form for the 2-halo term then becomes
\begin{align}
\label{eq:dmpower2a}
    P_{\rm DM}^{2h}(k,r) =  \int\int & I_{\rm DM}(k,m_1) I_{\rm DM}(k,m_2) \nonumber \\
    &\times b(m_1,r)b(m_2,r)P_{m}(k) dm_2 dm_1
\end{align}
This is separable, so long as the integration limits are not inter-dependent (this is not the case in general, cf. \S\ref{sec:theory:exclusion}), and in this simplest case (where we do not attempt to exclude correlations within halos) we have
\begin{equation}
    \label{eq:dmpower2}
    P_{DM}^{2h}(k,r) = P_m(k) \left[ \int  I_\text{DM}(k,m) b(m,r) dm \right]^2.
\end{equation}

Note that all of these expressions denote the 2-halo term as a function of both $k$ and $r$ (as opposed to the 1-halo term which is just a function of $k$). 
This arises either due to the appropriate limits of the integration being dependent on the real-space size of halos, or because the halo bias may be scale-dependent. 
In cases when either of these is true, to recover the power spectrum as a function of $k$ only, we Hankel-transform to $\xi(r)$ and then back to $P(k)$.

@@

The CDM transfer function has been modelled extensively in the literature. 
A number of functional forms have been proposed \citep[][hereafter EH]{Bond1984,Bardeen1986,Eisenstein1998}, along with a series of \textit{Boltzmann} codes, which calculate the transfer function from first principles with perturbation theory techniques \citep{Zaldarriaga2000,Lewis2000,Blas2011}.

Figure \ref{fig:power} shows the typical shape of the CDM linear power spectrum, while highlighting differences between forms for $T(k)$ found in the literature. Both small- and large-scales have small fluctuations, with a characteristic turnover at scales of $\sim 0.02h/$Mpc. The low-$k$ form asymptotes to $P(k \ll 0.02 ) \propto k^{n_s}$, while the high-$k$ behaviour is more complicated, not obeying a power-law relation. Only the Boltzmann code is able to fully capture the wiggles introduced by the baryon acoustic oscillations (BAOs), though the EH form including baryons closely follows up to $k \sim 2 h^{-1}{\rm Mpc}$. 
In fact, all curves (except that of \citet{Bond1984}) are within $\sim 10\%$ over the plotted range.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/transfer_models.pdf}
  \caption[Comparison of transfer function models]{A selection of transfer function models, shown as linear power spectra, including analytic approximations \citep{Bond1984,Bardeen1986,Eisenstein1998} and Boltzmann solutions. Details for each model (referenced to the alias given in the legend) can be found in Table \ref{tab:models_transfer}. The general shape is reproduced in each case, with details of baryon acoustic oscillations differing. In the bottom panel, we show the relative fraction compared to the CAMB transfer function, which renders all models without baryons as having oscillations on the acoustic scales.}
  \label{fig:power}
\end{figure}

The calculation of the halo-centre power spectrum is commonly approximated by Eq. \ref{eq:halo_centre_power}, i.e.
\begin{equation*}
    P_{hh}(k,r|m_1,m_2) \approx b(m_1,r)b(m_2,r)P_m(k),
\end{equation*}
 in which an estimate of the matter power spectrum is linearly biased. Typically, a \textit{nonlinear} estimate of the matter power is used, for which the corrections of \citet[][hereafter \textsc{halofit}]{Smith2003} are employed. We do not present the details here, except to mention that the modifications are applied as ratios of $P_\text{m,lin}(k)$, which are obtained via arguments motivated by the halo model. These modifications are expressed as a series of equations with coefficients fitted to numerical simulations. Updated coefficients were introduced in \citet{Takahashi2012}. Note also that the normalisation of $P_\text{m,nl}(k)$ is ident\-ical to $P_\text{m,lin}(k)$, and is calculated using the linear form, since $\sigma_8$ refers to the linear power spectrum. The effect of introducing non-linearities is to increase small-scale power, with the recent coefficient updates slightly enhancing this effect.

@@

The SO algorithm finds peaks in the density field and grows spheres around them until the contained density drops below a threshold density.
All particles within the final sphere are considered part of the halo. 
This definition is more useful in the context of observations (of clusters, for example) as it can be applied to the measured (projected) density fields\footnote{Note that the 3D SO definition is not \textit{immediately} applicable to observations without forward-modelling and scaling relations, but they are far more applicable than FoF masses.}.

@@

Having defined a spatial filter on the primordial density field, one can proceed to define the Lagrangian radius, $R(m)$, which is the radius of a spherical region of homogeneous space containing the mass $m$ (or more generally, the scale of a particular filter function required to contain the mass, cf. \S\ref{sec:halomod:components:window}). For the standard top-hat, this is
\begin{equation}
    R_L(m) = \left(\frac{3m}{4\pi\bar{\rho}_0}\right)^{1 / 3},
\end{equation}
with $\bar{\rho}_0$ the mean density of the Universe at the present epoch.

The ``mass variance'', $\sigma^2(R_L)$ is thus the variance on a scale from which the matter in halos is thought to have originated. Notably, the mass variance in spheres of Lagrangian radius 8 Mpc/$h$ yields the power spectrum normalisation, $\sigma_8$.

\subsection{Mass Function}
\label{sec:massfunction}
The mass function quantifies the number of dark matter halos per unit
mass per unit comoving volume of the Universe. 
%We do not include a detailed summary of the theory of the mass function here, and direct the reader to \citet{Murray2013a} for a review and table of fitting functions. 
The most commonly adopted theoretical formulation of the mass function is via the extended Press-Schechter (EPS) formalism \citep{Press1974,Bond1991}, in which density peaks form through gravitational instability and evolve into halos when they cross a density threshold, $\delta_c$. 
In this formalism, the mass function is a universal function of the peak-height, $\nu = \delta_c/\sigma$\footnote{Note that some authors use $\nu = \delta^2_c/\sigma^2(m)$, i.e. the square of our definition.}, which captures the variation from cosmological parameters and the redshift \citep[but note that several recent studies find significant non-universality with redshift, and have provided fits that explicitly account for this departure]{Jenkins2001}.

The mass function can be defined as
\begin{equation}
\label{eq:hmf}
  n(m) \equiv \frac{dn}{dm} = \frac{\rho_0}{m^2} f(\sigma) \left|\frac{d\ln\sigma}{d\ln m}\right|,
\end{equation}
or using the peak-height, as
\begin{equation}
    \label{eq:theory:hmf_nu}
    \frac{dn}{dm} = \frac{\rho_0}{m} f(\nu) \frac{d \nu}{d m}.
\end{equation}

For an Einstein-de Sitter universe (in which the matter density equals the critical density and there is no dark energy), $\delta_c \approx 1.686$, and its value is very weakly dependent on cosmology\footnote{Note that even small changes in $\delta_c$ can affect the high-mass end of the HMF significantly. However, many authors assume $\delta_c = 1.686$ regardless of cosmology and thus many fits of the HMF are defined under this assumption.}
The function $f(\sigma) \equiv \nu f(\nu)$, denoting the fraction of matter collapsed into halos of peak-height in a logarithmic bin around $\nu$, has been modelled extensively in the literature, and is generally fit to numerical simulations\footnote{Note that some authors use $f(\nu)$ to describe what we have defined as $\nu f(\nu)$.}.

It can be advantageous (but not strictly necessary) to choose a fitting function which is normalisable, ie.
\begin{equation}
    \label{eq:fnu1}
    \int f(\nu) d\nu = 1.
\end{equation}
Several fits in the literature obey this relation \citep[eg.][]{Sheth2001,Peacock2007,Tinker2010},  which ensures that all mass is bound in halos. This is particularly attractive for HM calculations, in which it is consistent with the basic assumptions of the framework.
